From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-10-01 paper.tex
Redacted for double-blind review

\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}
\newcommand{\etal}{\textit{et al.}\xspace}
\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\etc}{\textit{etc.}\xspace}
\newcommand{\vs}{\textit{vs.}\xspace}

@@

\title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}

@@

Engagement, however, is a complex, multi-faceted cognitive mechanism that is
only indirectly observable. This article formalizes \emph{with-me-ness} as one
of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
field of {\it Computer-Supported Collaborative Learning}, measures in a
well-defined way to what extend the human is \emph{with} the robot over the
course of an interactive task. As such, it is an meaningful precursor of
engagement.

@@

Building capable social agents requires to endow them with a range of perceptual
capabilities: however, while face and object tracking and recognition, path
planning, speech recognition or task learning are some of the current major
research directions, existing literature tends to evaluate each of these
algorithms in their own metric space, without considering the interaction
quality at a global level. Anzalone~\etal~\cite{anzalone} have recently argued
in favor of such a global evaluation, and they propose to assess these
algorithms in term of their capability to \emph{obtain the desired effect} in a
human-robot interaction context. They correspondingly propose metrics built
around the measurement of \emph{engagement} as indicator of the quality of the
experience.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating non-verbal cues (gaze and smiles) with the current
state of the interaction in a Bayesian model. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
\emph{connection} events such as directed gaze, mutual facial gaze is proposed.
Not relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture
as a determining factor for the assessment. Finally, a recent study with social
robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based
on non-verbal cues but they also underline possible limitations in long-term
scenarios.

@@

The following sections expose the full methodology, starting with the estimation
of attention: we present in the next section a novel method for on-line estimation
of the focus of attention based on fast 6D head pose estimation. We validate
this technique in section~\ref{sec:expe} with a real-world field study involving
children. Section~\ref{sec:withmeness} formally introduce the concept of
\emph{with-me-ness}. We present how to compute it over the course of the
interaction to eventually build a new \emph{in-the-moment} measurement of the
quality of interaction. We finally validate and discuss this metric by comparing
it to manual post-hoc annotations of the video-recordings of the interaction.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{head_pose}
    \caption{The 6D head pose is estimated by fitting a 3D model of an
        adult head (left) onto the detected 2D features of the face (right). We
        rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
        (three are depicted: the sellion -- the nasal depression --, the left
        tragion and the menton). The 3D origin of the head is set at the sellion.}

@@

As presented above, we use the 6D head pose as an approximation of the actual
gaze direction, and we further approximate from here the participant's field of
attention. The assumption that such an approximation of the field of attention
allows to derive the actual focus of attention needs to validated
experimentally.  Our proposed experiment involves child-robot interactions in
the context of handwriting remediation.  This section details the experimental
procedure and presents our results.

@@

The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
which tries to engage a child in handwriting tasks using a \emph{learning by
teaching} paradigm (\ie the child is the teacher, and he/she attempts to improve
the robot's handwriting). A tactile tablet is used as writing support.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\columnwidth]{realSetup}
    \caption{\small Picture of the interaction with one of the children.}
    \label{fig:realSetup}
\end{figure}

@@

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number, or a
full word), and presents a tactile tablet (equipped with a custom writing
application) to the robot. The robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pulls back the tablet, corrects the robot's attempt by writing on top of or
next to the robot's writing, and ``sends'' his/her demonstration to the robot by
pressing a small button on the tablet. The robot learns from this demonstration
and tries again. The child continues the turn-taking until they decide to train
the robot on another word. In total, the children performed on average 12.16
($SD=2.61$) rounds of writing (complete details on the rationale and
implementation of this experiment can be found in~\cite{Hood:2015}).

@@

The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent the possible point
of interests as individual frames: an object is considered to be in focus when
its frame lies within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

The implementation of the hand-writing activity itself has been presented by
Hood~\etal in~\cite{Hood:2015}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
        \emph{with-me-ness} over the whole $\approx$17min long interaction of
        subject~2. The top chart is similar to Figure~\ref{fig:realExpected}
        with the \emph{expected} attentional targets added in green. The bottom
        diagram represents the instantaneous level of with-me-ness over a
        sliding window of 30 seconds. The blue line is the with-me-ness as estimated
        by the robot, the orange line is the with-me-ness computed from manually
        annotated attentional targets.  Pearson's correlation between both
        series for this subject: $r(973)=0.58, p < 0.001$.}

@@

The experimental validation has been conducted with six children in face-to-face
interaction with an autonomous robot, over a total duration of about 2 hours. It
shows that {\it 1)} most of the time, we are able to estimate the head pose of
these children; {\it 2)} based on these head poses only, the instantaneous focus
of attention as computed by the robot does reach a good level of accuracy, with
however one inaccurate outlier out of the six participant; {\it 3)} the robot is
able to compute in real-time a level of \emph{with-me-ness} that correlates
strongly with the ground-truth.

@@

This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
National Science Foundation through the National Centre of Competence in
Research Robotics.

>>>>

\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}
\newcommand{\etal}{\textit{et al.}\xspace}
\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\etc}{\textit{etc.}\xspace}
\newcommand{\vs}{\textit{vs.}\xspace}
\newcommand{\redacted}{\textit{\sc [REDACTED FOR REVIEW]}\xspace}

@@

\title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}

%\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
\author{Redacted for double-blind review}

@@

Engagement, however, is a complex, multi-faceted cognitive mechanism that is
only indirectly observable. This article formalizes \emph{with-me-ness} as one
of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
field of {\it Computer-Supported Collaborative Learning}, measures in a
well-defined way to what extent the human is \emph{with} the robot over the
course of an interactive task. As such, it is a meaningful precursor of
engagement.

@@

Building capable social agents requires to endow them with a range of perceptual
capabilities: however, while face and object tracking and recognition, path
planning, speech recognition or task learning are some of the current major
research directions, existing literature tends to evaluate each of these
algorithms in their own metric space, without considering the interaction
quality at a global level. Anzalone~\etal~\cite{anzalone} have recently argued
in favor of such a global evaluation, and they propose to assess these
algorithms in terms of their capability to \emph{obtain the desired effect} in a
human-robot interaction context. They correspondingly propose metrics built
around the measurement of \emph{engagement} as indicator of the quality of the
experience.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating non-verbal cues (gaze and smiles) with the current
state of the interaction in a Bayesian model. While they report a high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
\emph{connection} events such as directed gaze, mutual facial gaze is proposed.
Not relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture
as a determining factor for the assessment. Finally, a recent study with social
robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based
on non-verbal cues but they also underline possible limitations in long-term
scenarios.

@@

The following sections expose the full methodology, starting with the estimation
of attention: we present in the next section a novel method for on-line estimation
of the focus of attention based on fast 6D head pose estimation. We validate
this technique in section~\ref{sec:expe} with a real-world field study involving
children. Section~\ref{sec:withmeness} formally introduces the concept of
\emph{with-me-ness}. We present how to compute it over the course of the
interaction to eventually build a new \emph{in-the-moment} measurement of the
quality of interaction. We finally validate and discuss this metric by comparing
it to manual post-hoc annotations of the video-recordings of the interaction.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{head_pose}
    \caption{The 6D head pose is estimated by fitting a 3D model of an
        adult head (left) {\it\sc[FACE REDACTED FOR REVIEW]} onto the detected 2D features of the face (right). We
        rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
        (three are depicted: the sellion -- the nasal depression --, the left
        tragion and the menton). The 3D origin of the head is set at the sellion.}

@@

As presented above, we use the 6D head pose as an approximation of the actual
gaze direction, and we further approximate from here the participant's field of
attention. The assumption that such an approximation of the field of attention
allows to derive the actual focus of attention needs to be validated
experimentally.  Our proposed experiment involves child-robot interactions in
the context of handwriting remediation.  This section details the experimental
procedure and presents our results.

@@

The experiment, part of the \redacted project, involves a robot
which tries to engage a child in handwriting tasks using a \emph{learning by
teaching} paradigm (\ie the child is the teacher, and he/she attempts to improve
the robot's handwriting). A tactile tablet is used as writing support.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\columnwidth]{realSetup}
    \caption{\small Picture of the interaction with one of the children
    {\it\sc[FACE REDACTED FOR REVIEW]}.}
    \label{fig:realSetup}
\end{figure}

@@

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number, or a
full word), and presents a tactile tablet (equipped with a custom writing
application) to the robot. The robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pulls back the tablet, corrects the robot's attempt by writing on top of or
next to the robot's writing, and ``sends'' his/her demonstration to the robot by
pressing a small button on the tablet. The robot learns from this demonstration
and tries again. The child continues the turn-taking until they decide to train
the robot on another word. In total, the children performed on average 12.16
($SD=2.61$) rounds of writing (complete details on the rationale and
implementation of this experiment can be found in \redacted).

@@

The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent the possible point
of interests as individual frames: an object is considered to be in focus when
its frame lies within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\redacted.

The implementation of the hand-writing activity itself has been presented by
\redacted.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
        \emph{with-me-ness} over the whole $\approx$17min long interaction of
        subject~2. The top chart is similar to Figure~\ref{fig:realExpected}
        with the \emph{expected} attentional targets added in green. The bottom
        diagram represents the instantaneous level of with-me-ness over a
        sliding window of 30 seconds. The blue line is the with-me-ness as estimated
        by the robot, the orange line is the with-me-ness computed from manually
        annotated attentional targets.  Pearson's correlation between both
        series for this subject: $r(973)=0.58, p < .001$.}

@@

The experimental validation has been conducted with six children in face-to-face
interaction with an autonomous robot, over a total duration of about 2 hours. It
shows that {\it 1)} most of the time, we are able to estimate the head pose of
these children; {\it 2)} based on these head poses only, the instantaneous focus
of attention as computed by the robot does reach a good level of accuracy, with
however one inaccurate outlier out of the six participants; {\it 3)} the robot is
able to compute in real-time a level of \emph{with-me-ness} that correlates
strongly with the ground-truth.

@@

\redacted
